---
title: "Candidate Debt Exploration"
author: "Divya Kukati, Jason Tuenge, and Steve Carpenter"
date: "25 September 2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

As members of the campaign committee for an upcoming election, we would be interested in knowing
as much as possible about debt reported by candidates in previous elections. Using the monthly
voter registration statistics for registered voters in Oregon from 2012, we performed an 
exploratory analysis to help the agency address their goals. Our objectie was to understand 
how campaign characteristics are related to the candidates debt.

We Were given a dataset on voter registration from 2012 **"CandidateDebt.csv"**.  The dependent 
(or target) variable in this data is named "amount". We were told that the labels of the variables
would be as listed below, and that the rest of the variables should be self-explanatory

**id**: internal identifier that corresponds to a single expenditure record. 
**reportnumber**: identifier used for tracking the individual form 
**origin**: This field shows from which filed report-type the data originates. 
**filerid**: The unique id assigned to a candidate. 
**filertype**: Indicates if this record is for a candidate.
**filername**: The candidate or committee name as reported on the candidates registration. 
**firstname**: This field represents the first name, as reported by the filer. 
**middleinitial**: This field represents the middle initial, as reported by the filer. 
**lastname**: This field represents the last name, as reported by the filer.
**office**: The office sought by the candidate.
**legislativedistrict**: The Washington State legislative district. 
**position**: The position associated with an office. 
**party**: The political party as declared by the candidate on their registration.
**jurisdiction**: The political jurisdiction associated with the office of a candidate.
**jurisdictioncounty**: The county associated with the jurisdiction of a candidate. 
**jurisdictiontype**: The type of jurisdiction this office is: Statewide, Local, etc.
**electionyear**: The election year in the case of candidates.
**amount**: The amount of the debt incurred or order placed.
**recordtype**: This field designates the item as a debt.
**fromdate**: The start date of the period for the report on which this debt record was reported.
**thrudate**: The end date of the period for the report on which this debt record was reported.
**debtdate**: The date that the debt was incurred.
**code**: The type of debt. 
**description**: The reported description of the transaction.
**vendorname**: The name of the vendor or recipients name. 
**vendoraddress**: The street address of the vendor or recipient.
**vendorcity**: The city of the vendor or recipient.
**vendorstate**: The state of the vendor or recipient.
**vendorzip**: The zip code of the vendor or recipient.


## Introduction

The research question that motivated our analysis was: 
How are campaign characteristics related to the candidates debt?

## Initial File Analysis
### Describe your data set. What types of variables does it contain? How many observations are there?

The Data is in a file called "CandidateDebt.csv" which has 28 variables and 1043 observations. 
The variable **reportnumber** was of integer class, and the rest were factors with 1-141 levels. 
All were listed above (i.e., in the assignment), and no 'id' variable was present.

```{r car}
library(car)
library(ggplot2)
par(oma= c(2,1,5,2) + 0.1) #Sets outside margins : b, l, t, r 
par(mar = c(4,1,5,1)) #Sets plot margins
cdebt_original = data.frame(read.csv("CandidateDebt.csv"))
str(cdebt_original)
```


### Evaluate the data quality. Are there any issues with the data? Explain how you handled these potential 
###issues. Explain whether any data processing or preparation is required for your data set.

It is already apparent that:

* As we would hope, each **reportnumber** is unique. 
* Similarly, there are as many unique **filernames** as unique 
  [**lastname**, **firstname**, **middlename**] triples.
* There are more than 7 **reportnumbers** per **filerid** on average. Consequently, a given person will be 
  counted multiple times if corresponding rows aren't consolidated into a total or set of totals. 
* There are 7 more levels for **filerid** than for **filername**. In all cases, simple internet 
  searches revealed that candidates had held one office going into the campaign and ran for a different 
  office for the 2012 election (**filername** in "FROCKT DAVID S", "KELLEY TROY X", "MCKENNA ROBERT M",
  "PROBST TIMOTHY P", "WILLIAMS BRENDAN W"), held one office and been appointed another mid-term and 
  by the 2012 election ran for the appointed office (**filername** == "ROLFES CHRISTINE N"), 
  or held no office and started running for one office but ultimately changed their campaign to 
  run for a different office at the end of the election (**filername** == "LADENBURG JOHN W SR").

References:

* https://ballotpedia.org/Dave_Frockt
* https://en.wikipedia.org/wiki/Troy_Kelley
* http://bit.ly/2fnAH9a
* https://en.wikipedia.org/wiki/Rob_McKenna
* https://en.wikipedia.org/wiki/Christine_Rolfes
* https://ballotpedia.org/Tim_Probst
* https://ballotpedia.org/Brendan_Williams

```{r uniqueValues}
length(cdebt_original$reportnumber) - length(unique(cdebt_original$reportnumber))

comb_fmlnames <- length(sort(unique(paste(cdebt_original$lastname, 
                                          cdebt_original$firstname, 
                                          cdebt_original$middleinitial))))
#shared lastname: FARRELL, JOHNSON, LADENBURG, MORRIS, THOMAS
length(unique(cdebt_original$filername)) - comb_fmlnames 

length(unique(cdebt_original$reportnumber)) / length(unique(cdebt_original$filerid))

length(unique(cdebt_original$filerid)) - length(unique(cdebt_original$filername))
```

When we reviewed the file we found what seemed to be an undocumented extra column right after the **office** 
column that was similar in nature but it offset the headers for the rest of the columns. 
For instance, the **legislativedistrict** data was now under the **position** header, the **party** 
data was under the **jurisdiction** header, etc... For example, 'DEMOCRAT' should refer to party 
rather than to jurisdiction. We verified that the **office** column was setup correctly by selecting 
multiple records and doing an internet search to see what office the specific person (**filername**) 
was campaigning for in Washington State in 2012. For instance, Sheryl McCloud (**filername** == "MCCLOUD SHERYL G") 
ran for State Supreme Court Justice and won the seat in 2012. The data in the extraneous columnn indicated 
that she is somehow linked to a State Representative seat; we couldn't find any evidence of this online.

References:

    * https://ballotpedia.org/Sheryl_McCloud
    * http://sdc.wastateleg.org/frockt/
    * https://ballotpedia.org/Washington_elections,_2012

We decided to relabel this phantom column to **office2**. In addition, we observed that a number of 
fields were empty or contained '#N/A' text. We converted these to 'NA' while loading the repaired CSV file, 
and found that this resulted in several columns (**legislativedistrict**, **position**, **electionyear**) 
changing from a factor type to an integer type. Similarly, the key **amount** variable changed from a 
factor type to a numeric type, thereby enabling quantitative analysis of debt. 

```{r revisedCSV}
colNames <- c("reportnumber", "origin", "filerid", "filertype", "filername", "firstname", "middleinitial",
              "lastname", "office", "office2", "legislativedistrict", "position", "party",
              "jurisdiction", "jurisdictioncounty", "jurisdictiontype", "electionyear", 
              "amount", "recordtype", "fromdate", "thrudate", "debtdate", "code",
              "description", "vendorname", "vendoraddress", "vendorcity", "vendorstate")
cdebt_revised_interim <- read.csv("CandidateDebt.csv", col.names = colNames, stringsAsFactors = T,
                         na.strings = c(NA,"NA","#N/A" ,"", " "))
str(cdebt_revised_interim)
```

We then removed the 56 rows containing NA values in the **amount** column, as these could not be 
quantitatively related to debt. In fact, these rows were mostly comprised of NA values, 
thereby confounding comparison with other variables as well.

```{r amountNAs}
cdebt_revised = subset(cdebt_revised_interim,  !is.na(amount))
#cdebt_revised = subset(cdebt_revised_interim,  !is.na(amount) & jurisdiction != "LEG DISTRICT 01 - SENATE")
#str(cdebt_revised)
length(cdebt_original$amount) - length(cdebt_revised$amount)
```

A number of variables were still substantially comprised of NA values.
As is typical in data analysis, since these values were not impacting our ability 
to explore the dataset, we decided to keep them in the analysis and handle 
any evaluations that are impacted by NA values.

```{r origin}
paste(c(round(1 - length(na.omit(cdebt_revised$legislativedistrict)) / length(cdebt_revised$legislativedistrict), 3), "(legislativedistrict)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$position)) / length(cdebt_revised$position), 3), "(position)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$jurisdictioncounty)) / length(cdebt_revised$jurisdictioncounty), 3), "(jurisdictioncounty)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$code)) / length(cdebt_revised$code), 3), "(code)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$description)) / length(cdebt_revised$description), 3), "(description)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$vendoraddress)) / length(cdebt_revised$vendoraddress), 3), "(vendoraddress)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$vendorcity)) / length(cdebt_revised$vendorcity), 3), "(vendorcity)"), collapse = " ")
paste(c(round(1 - length(na.omit(cdebt_revised$vendorstate)) / length(cdebt_revised$vendorstate), 3), "(vendorstate)"), collapse = " ")
```

The MM/DD/YY format of the 3 date variables (**fromdate**, **thrudate**, **debtdate**)
led to their factor classification. To obtain useful data, We coerced from factor to date.

```{r dates}
#summary(cdebt_revised$debtdate)
formattedFromdate <- as.Date(as.character(cdebt_revised$fromdate), format = "%m/%d/%y")
formattedThrudate <- as.Date(as.character(cdebt_revised$thrudate), format = "%m/%d/%y")
formattedDebtdate <- as.Date(as.character(cdebt_revised$debtdate), format = "%m/%d/%y")
summary(formattedDebtdate)
``` 

Following are some additional observations made while repairing the CSV file:

* All records associated with **vendorname** == "HICKEY GAYLE" are labeled as **party** == "REPUBLICAN" 
  regardless of the party of the candidate in public records and in the data set for other debt transactions. 
  This has the effect that analysis on party will show records for some candidates as both DEMOCRAT and REPUBLICAN. 
  This is incorrect and should be fixed based on the candidate's official party alignment. 
  We did some analysis and in no cases did the sample we selected switch from DEMOCRAT to REPUBLICAN or vice versa.
  
* Both **party** == "INDEPENDENT" records in the data set are also affiliated 
  with a single **vendorname** == "JEFFREYS TERRI M" who just so happens to be a candidate for 
  Mason County Commissioner affiliated as an INDEPENDENT. In some records she is labelled as a DEMOCRAT, 
  in others as a REPUBLICAN (but only in the "HICKEY GAYLE" scenario identified above), 
  and with regard to Postage as an INDEPENDENT. Because the new sources label her as an INDEPENDENT and 
  most of the records label her as a DEMOCRAT, we will not change the party affiliation. 
  The other candidate who used the **vendorname** == "JEFFREYS TERRI M" was a DEMOCRAT named David Frockt. 
  It is assumed he likely shared the cost of postage with Ms Jeffreys and the record was labeled incorrectly.
  
* Whereas **office** is related to filername, five variables 
  (**office2**, **legislativedistrict**, **jurisdiction**, **jurisdictioncounty**, and **jurisdictiontype**) 
  are all related to one another but do not appear to be related to **office** or **filername**.
  For example, Terri Jeffreys (http://www.terrijeffreys.org/2012/06/19/citizens-to-elect-terri-jeffreys/) 
  ran for county commissioner (**office**) rather than governor (**office2**), 
  and the corresponding jurisdiction ('GOVERNOR, OFFICE OF') would pertain to the latter. 
  The relevance of these 5 variables is unclear. 

```{r vendors}
hic <- sum(cdebt_revised$vendorname == "HICKEY GAYLE")
hicrep <- sum((cdebt_revised$vendorname == "HICKEY GAYLE") & 
              (cdebt_revised$party == "REPUBLICAN"))
hic - hicrep

sum(cdebt_revised$vendorname == "JEFFREYS TERRI M")
sub_tjven <- subset(cdebt_revised, vendorname ==  "JEFFREYS TERRI M",
                    c(filername, party, vendorname))
sub_tjven
sub_tjdfc <- subset(cdebt_revised, ((filername == "JEFFREYS TERRI M") |
                                    (filername == "FROCKT DAVID S")),
                    c(filername, party, vendorname))
sub_tjdfc[order(sub_tjdfc$filername),]

sub_tjoff <- subset(cdebt_revised, (filername == "JEFFREYS TERRI M"),
                    c(filername, office, office2))
sub_tjoff
```

Last, the dataset contains several variables that appear unlikely to contribute much or any information:

  * **firstname**, **middleinitial**, **lastname**: These are redundant to **filername**, as noted above.
  * **origin**, **filertype**, **recordtype**: Factor has only one level.
    Some rows in **recordtype** are NA, but in these cases the entire row is NA.
  * **electionyear**: All values are either either 2012 or NA.

```{r uniform}
summary(cdebt_revised$origin)
summary(cdebt_revised$filertype)
summary(cdebt_revised$recordtype)
summary(cdebt_revised$electionyear)
```


## Univariate Analysis of Key Variables: Use visualizations and descriptive statistics to perform a univariate analysis of each key variable. Be sure to describe any anomalies, coding issues, or potentially erroneous values. Explain how you respond to each issue you identify. Note any features that appear relevant to statistical analysis. Discuss what transformations may be appropriate for each variable.

The 'amount' values are all positive and range from around 3 to 19,000 
(units are unclear but are presumably dollars or perhaps thousands of dollars), 
with most values below 1,000. The mean is more than 4x the median, 
and a histogram confirms the distribution is positively skewed. 

```{r amount}
amount = cdebt_revised$amount
summary(amount)
hist(cdebt_revised$amount, breaks = seq(0, 20000, by = 1000), main = "Histogram of the 'amount' variable",  border="black", col="#00A5DF", xlab = "Amount of debt" ,ylab="Number of debts", ylim = c(0, 800))
```

Nearly 65% of filings (by count rather than amount) were by 
Democrats -- more than twice the number filed by Republicans.  

```{r party}
party <- table(cdebt_revised$party)
round(summary(cdebt_revised$party) / length(cdebt_revised$party), 3)
par(las=1, mar=c(5,10,2,1)) #Sets outside margins : b, l, t, r 
colorPalette <- c("#E69F00", "#009E73", "#FF0000", "#0000FF") # I , N , R , D
barplot(party[order(party)], horiz = TRUE , las = 1, main ="Number of Debts by Party" , xlab="No of Debts" ,cex.axis = .9 ,cex.names= 0.9 , col = colorPalette , xlim = c(0, 700))
```

Most debt entries (nearly 54% of total) were filed by state representatives. 
State senators had the next-largest share, with 12% of the total count.

```{r office}
round(summary(cdebt_revised$office) / length(cdebt_revised$office), 3)
office <- table(cdebt_revised$office)
par(las=1, mar=c(5,8,2, 1)) #Sets plot margins : b, l, t, r 
barplot(office[order(office)], horiz = TRUE , las = 1, main ="Number of Debts by Office", xlab="No of Debts" ,cex.names=0.5 , col = terrain.colors(16),xlim = c(0, 600)) 
```

Nearly 47% of entries are associated with vendorcity of Seattle, 
followed by Woodinville and Washington, D.C.

```{r vendorcity}
count_Seattle = subset(cdebt_revised, vendorcity == "SEATTLE")
count_cities = subset(cdebt_revised, !is.na(vendorcity))
round(length(count_Seattle$vendorcity) / length(count_cities$vendorcity),3)
#vendorcity <- table(cdebt_revised$vendorcity)
#par(las=1, mar=c(2.5,8,2, 0.8)) #Sets plot margins : b, l, t, r 
#barplot(vendorcity[order(vendorcity)], horiz = TRUE , las = 1, main ="Number of debts by vendorcity", xlab="Number of debts" ,cex.names=0.5 , col = heat.colors(29) , xlim = c(0, 500)) 
```

Fully 25% of entries have "RE-ORDER TEE SHIRTS" in the description field. 
This factor level appears more frequently than the next three levels 
(CONSULTING/TRAVEL, ACCOUNTING/COMPLIANCE, NOVEMBER TREASURY) combined. 

```{r description}
count_shirts = subset(cdebt_revised, description == "RE-ORDER TEE SHIRTS")
count_descriptions = subset(cdebt_revised, !is.na(description))
round(length(count_shirts$description) / length(count_descriptions$description),3)

# Top 15 entries 
description <- table(cdebt_revised$description)
sortedDescription = description[order(description, decreasing = TRUE)]
#sortedDescription[15:1]

par(las=1, mar=c(2.5,10,2, 0.7)) #Sets plot margins : b, l, t, r 
barplot(sortedDescription[15:1], horiz = TRUE , las = 1, main ="Number of debts by description\n for top 15 entries", xlab="Number of debts" ,cex.names=0.6 , col = terrain.colors(16) , xlim = c(0, 300) , width = 10) 

```

## Analysis of Key Relationships: Explore how your outcome variable is related to the other variables in your dataset. Make sure to use visualizations to understand the nature of each bivariate relationship. What tranformations can you apply to clarify the relationships you see in the data? Be sure to justify each transformation you use.

### Debt Date

Although **amount** at first glance appeared to increase with **debtdate**, 
there was no real correlation.  We also observed that there seems to be a 
surge in debts in the middle of the reportnumber range, presumably due to values 
for this variable increasing with time (i.e., probably issued sequentially).

```{r debtdate}
#plot(cdebt_revised$reportnumber, amount)
#abline(lm(amount ~ cdebt_revised$reportnumber))
plot(formattedDebtdate, amount, main = "Debt 'amount' with respect to 'debtdate'", xlab = "debtdate")
abline(lm(amount ~ formattedDebtdate))
```

### Amount by Party

Most of the debt was filed by Democrats -- nearly 10x the amount filed by Republicans. 
The aforementioned 2x difference in filing count does not explain this gap between parties.

```{r totalAmountVsParty}
amount_sum_party = tapply(cdebt_revised$amount, cdebt_revised$party, FUN=sum )
(amount_sumK_party = amount_sum_party)
# Still suggest having /1000 for easy reading and setting xlim

par(las=1, mar=c(5,10,2,0.8))
colorTAPalette <- c( "#E69F00", "#009E73","#FF0000","#0000FF") # N , I , D , R
barplot(amount_sumK_party[order(amount_sumK_party)], horiz = TRUE , las = 1, main ="Total Debt Amount by Party" , xlab="Total debt amount" ,cex.axis = .9 ,cex.names= 0.9 , col = colorTAPalette)
```

If there were more Democratic candidates than Republican candidates, 
it wouldn't necessarily be surprising to see Democrats spending more as a party than Republicans. 
To get a better sense of typical spending spending by candidates of each party, we plot the average. 

```{r meanAmountVsParty}
amount_means_party = by(cdebt_revised$amount, cdebt_revised$party,  mean)
amount_means_party[order(amount_means_party, decreasing = TRUE)]

par(las=1, mar=c(5,10,2,0.8)) #Sets outside margins : b, l, t, r 
colorMAPalette <- c( "#E69F00", "#FF0000","#009E73","#0000FF") # I , N , R , D

barplot(amount_means_party[order(amount_means_party)], horiz = TRUE , las = 1, main ="Mean Debt Amount by Party" , xlab="Mean debt amount" ,cex.axis = .9 ,cex.names= 0.9 , col = colorMAPalette, xlim = c(0, 2000)) 
```

A boxplot gives a better sense of the distribution, and provides the following insights:
* The distribution is quite positively skewed for Democrats, with outliers yielding an 
  average that is more than 4x the median.
* Similarly, an outlier results in substantial positive skew for non-partisan candidates.
* The numerous 283.25 amounts in the Republican data generate a median of this value, 
  with no interquartile range and multiple "outliers" that yields a relatively mild 
  (in comparison to Democrats) positive skew in the distribution. 

```{r medianAmountVsParty}
by(cdebt_revised$amount, cdebt_revised$party, median)

par(las=1, mar=c(5,10,2,0.3)) #Sets outside margins : b, l, t, r 
colorBAPalette <- c("#0000FF", "#E69F00", "#009E73", "#FF0000" ) # D , R , N , I <==> D , N , I , R
boxplot(cdebt_revised$amount ~ party, cex.axis = .6, horizontal = TRUE ,data = cdebt_revised,
main = "Debt amount by party", col = colorBAPalette, xlab = "Debt amounts")
```

### Amount by Office

Turning our attention to the relationship between office and debt amount, 
we see that state representatives filed 4x the amount filed by the second 
and third-highest offices (county commissioner and state senator, respectively).

```{r amount_total_office}
amount_sum_office = tapply(cdebt_revised$amount, cdebt_revised$office, FUN=sum)
options(scipen=5)
(amount_sumK_office = amount_sum_office)
# Still suggest having /1000 for easy reading and setting xlim
```

```{r totalAmountVsOffice}
par(las=1, mar=c(5,10,2,0.8)) #Sets outside margins : b, l, t, r 
barplot(amount_sumK_office[order(amount_sumK_office)], horiz = TRUE , las = 1, main ="Total Debt Amount by Office" , xlab="Total debt amount" ,cex.axis = .8 ,cex.names= 0.6 , col = terrain.colors(16) , xlim=c(0, 800000) )
```

Comparing with mean values as before paints an entirely different picture: 
We see that appeals court judges appear to have relatively high debt on an individual basis, 
and state representatives drop toward the lower end of the scale. 

```{r meanAmount_office}
amount_means_office = by(cdebt_revised$amount, cdebt_revised$office,  mean, na.rm = TRUE)
amount_means_office[order(amount_means_office, decreasing = TRUE)]
```

```{r meanAmountVsOffice}
par(las=1, mar=c(5,10,6,0.8)) #Sets outside margins : b, l, t, r 

barplot(amount_means_office[order(amount_means_office)], horiz = TRUE , las = 1, main ="Mean Debt Amount by Office" , xlab="Mean debt amount" ,cex.axis = .9 ,cex.names= 0.5 , col = terrain.colors(16), xlim = c(0, 4000)) 
```

A boxplot again reveals considerable positive skew for all offices excepting appeals court judges, 
which had no outliers but still exhibited substantial positive skew. 
Furthermore, the median was around 300 for all offices except appeals court judges, which had a median of nearly 10x this value. 

```{r amount_medians_Office}
amount_medians_office = by(cdebt_revised$amount, cdebt_revised$office, median, na.rm = TRUE)
amount_medians_office[order(amount_medians_office, decreasing = TRUE)]
```

```{r medianAmountVsOffice}
par(las=1, mar=c(3,10,2,0.3)) #Sets outside margins : b, l, t, r 
boxplot(cdebt_revised$amount ~ office, data = cdebt_revised, cex.axis = .6, horizontal = TRUE ,
main = "Debt Amount by Office" , las=2 , col = "#DDA266")

```

### Amount by Description

```{r amount_mean_Description}

amount_means_description = by(cdebt_revised$amount, cdebt_revised$description,  mean, na.rm = TRUE)
sortedMeans = amount_means_description[order(amount_means_description, decreasing = TRUE)]
sortedMeans[15:1]
```

```{r meanAmountVsDescription}
par(las=1, mar=c(3,12,2,1))#Sets outside margins : b, l, t, r 

barplot(sortedMeans[15:1], horiz = TRUE , las = 1, main ="Mean Debt by Office" , xlab="Mean Debt in $" ,cex.axis = .9 ,cex.names= 0.4 , col = terrain.colors(16), xlim = c(0, 20000)) 
```

Median for Consulting is pulled away from the mean which indicates that the distribution is negatively skewed

```{r amount_median_Description}
amount_median_description = by(cdebt_revised$amount, cdebt_revised$description,  median, na.rm = TRUE)
sortedMedian = amount_median_description[order(amount_median_description, decreasing = TRUE)]
sortedMedian[15:1]
```


```{r medianAmountVsDesc}
par(las=1, mar=c(3,12,2,1))#Sets outside margins : b, l, t, r 
barplot(sortedMedian[15:1], horiz = TRUE , las = 1, main ="Median Debt by Office" , xlab="Mean Debt in $" ,cex.axis = .9 ,cex.names= 0.4 , col = terrain.colors(16), xlim = c(0, 20000)) 

```
### Amount, Office, and Party

The below plot is an aggregate of Amount by Office and Party.
This plot shows that Candidates the same office but from different parties have different 
average amount of debts
For instance : State representative from Democrats has average debt of 1710 $ where as 
               State rep from Republicans has average debt of 443 $
Another ex   : State Senator from Non Partisans have average debt of  2624 $ whereas
               State Senator from Democrats have average debt of 1623$
               which contadricts the fact that average amount by Office.

```{r amountVsofficeVsParty}

amount_office_party <- aggregate(x = cdebt_revised$amount, by = cdebt_revised[c("office", "party")], FUN = mean, na.rm=TRUE)
amount_office_party

#Sets outside margins : b, l, t, r 
par(las=1, mar=c(3,10,2,0.3))

ggplot(data=amount_office_party, aes(x=party, y=x, fill=office)) +
    geom_bar(stat="identity", position=position_dodge())

```

## Analysis of Secondary Effects: What secondary variables might have confounding effects on the relationships you have identified? Explain how these variables affect your understanding of the data.

1. cdebt_revised = subset(cdebt_revised_interim,  !is.na(amount) & jurisdiction != "LEG DISTRICT 01 - SENATE")
2. We looked at deleting the jurisdiction related variables
3. We looked at deleting the values of leg district 01 - senate pertaining to Re-Order T-shirts entries, but that eliminate the  
   multiple entries

## Conclusion: Summarize your exploratory analysis. What can you conclude based on your analysis?

1) Real-World Data is very messy and can be influenced by:

* data collection processes
* metadata/data dictionary completeness
* willful instructor sabotage
* formatting for an audience/medium
* encoding techniques
* real-life scenarios (candidates changing their minds)
* analysis of originally unintended scenarios

2) Real-World Data Analysis is more meaningful if you apply domain knowledge 
  to the data set and use whatever sources you have at your disposal (internet, industry knowledge, etc...) 
  to understand what you are seeing; a good example is the inordinately large number of candidates
  that seemed to switch parties between Democrat and Republican. This happens in real-life,
  but based on anecdotal evidence (domain kknowledge) it doesn't happen very often. 
  So when it seems like it is happening more than expected, digging in to understand what is 
  really happening is critical. (eg., Gayle Hickey Tee-Shirt Re-orders are all Republican)

